---
title: "THSS 730 - Lab 4"
subtitle: "Confidence Intervals and Hypothesis Testing"

format: 
  html: 
    toc: true
    df-print: paged
    embed-resources: true
---

```{r, message=FALSE}
# Necessary libraries
library(tidyverse)
library(readxl)
library(glue)
```


## Chi-Square Distribution

If $X \sim N(\mu, \sigma^2)$, we may need to consider the sampling variance of $X$, which is often denoted as $S^2$. It turns out this statistic follows a random distribution: $$S^2 \sim \frac{\sigma^2 \chi^2_{n-1}}{n-1}$$

Note I'm using $S$ instead of lower case $s$ because it's common to use capital letters when denoting a random variable and lower case letters when denoting an observed value of that random variable. 

The chi-square ($\chi^2$) distribution:

- Has one parameter: degrees of freedom (df)
- Only takes on positive values and is always skewed to the right 
- The skewness diminishes as the degrees of freedom increase


In R, we can calculate probabilities and percentiles of the chi-square distribution using `pchisq()` and `qchisq()` functions.

```{r}

# Cumulative probability
pchisq(1, df = 9)

# Upper tail probability
1 - pchisq(3, df = 9)

# Percentiles (quantiles)
qchisq(0.975, df = 9)
qchisq(0.025, df = 9)

```

## F Distribution

If you have two independent samples from a $N(\mu_1, \sigma_1^2)$ and $N(\mu_2, \sigma_2^2)$ distribution, respectively, then under $H_0: \sigma_1^2 = \sigma_2^2$, $$\frac{S_1^2}{S_2^2} \sim F_{(n_1 - 1, n_2 - 1)}$$


The F distribution:
-	Has two parameters: numerator degrees of freedom (ndf) and denominator degrees of freedom (ddf)
-	Only takes on positive values and is generally skewed to the right 
-	The skewness depends on the relative magnitudes of the two degrees of freedom


For the F distribution, we use `pf()` and `qf()` functions in R.

```{r}
# Cumulative probability for F distribution
pf(1, df1 = 19, df2 = 9)

# Upper tail probability for F distribution
pf(1, df1 = 19, df2 = 9, lower.tail = FALSE)

# Percentiles (quantiles)
qf(0.975, df1 = 19, df2 = 9)
qf(0.025, df1 = 19, df2 = 9)

```

## Plotting Probability Density Functions (PDFs)

We can use `ggplot2` to plot the PDFs for the chi-square and F distributions for different degrees of freedom.

### Chi-Square Distribution Plot

```{r}

# Create data for chi-square PDF
x <- seq(0, 40, by  = 0.1)
df5 <- dchisq(x, df = 5)
df10 <- dchisq(x, df = 10)
df20 <- dchisq(x, df = 20)

chi_data <- data.frame(
  x = rep(x, 3),
  density = c(df5, df10, df20),
  df = factor(rep(c(5, 10, 20), each = length(x)))
)

ggplot(chi_data, aes(x = x, y = density, color = df)) +
  geom_line() +
  labs(title = "Chi-Square Distribution", x = "x", y = "Density") +
  theme_minimal() + 
  labs(x = "x value", y = "Density")
```

### F-Distribution Plot

```{r}
# Create data for F-distribution PDF

x <- seq(0, 10, by = 0.05)
f5_5 <- df(x, df1 = 5, df2 = 5)
f10_5 <- df(x, df1 = 10, df2 = 5)
f20_10 <- df(x, df1 = 20, df2 = 10)

f_data <- data.frame(
  x = rep(x, 3),
  density = c(f5_5, f10_5, f20_10),
  df = factor(rep(c("5 ndf, 5 ddf", 
                    "10 ndf, 5 ddf", 
                    "20 ndf, 10 ddf"), 
                  each = length(x)))
) %>% 
  mutate(df = fct_inorder(df))

ggplot(f_data, aes(x = x, y = density, color = df)) +
  geom_line() +
  labs(title = "F Distribution", x = "x", y = "Density") +
  theme_minimal()
```

## Confidence intervals

A 100% x (1-$\alpha$) confidence interval for $\sigma^2$  is given by:

$$\left[ \frac{(n-1)s^2}{\chi^2_{n-1,1-\alpha/2}}, \frac{(n-1)s^2}{\chi^2_{n-1,\alpha/2}} \right]$$

To obtain a $100\% \times (1 - \alpha)$ confidence interval for $\sigma$, take the square root of the confidence limits.  

**NOTE:** These confidence intervals for $\sigma^2$ and $\sigma$ are only valid for normally distributed data.

An approximate $100\% \times (1 - \alpha)$ confidence interval for the binomial parameter $p$ based on the normal approximation to the binomial distribution is given by:

$$\hat{p} \pm z_{(1 - \alpha/2)} \sqrt{\frac{\hat{p} \cdot \hat{q}}{n}}$$

or equivalently,

$$\left( \hat{p} - z_{(1 - \alpha/2)} \sqrt{\frac{\hat{p} \cdot\hat{q}}{n}}, \hat{p} + z_{(1 - \alpha/2)} \sqrt{\frac{\hat{p} \cdot\hat{q}}{n}} \right)$$

This is sometimes called the Wald method and is appropriate when $n \cdot \hat{p} \cdot \hat{q} \geq 5$.

When the normal approximation is not appropriate, exact binomial probabilities can be used to construct an exact confidence interval for p. This is sometimes called the Clopper-Pearson method. 


Going back now to the “lowbwt.xlsx” example. Recall that the data set has the following variables:
	
- `sbp` = systolic blood pressure
- `sex` = sex (1=male; 0=female)
- `tox` = toxemia diagnosis for mother (1=yes; 0=no)
- `grmhem` = germinal matrix hemorrhage (1=yes; 0=no)
- `gestage` = gestational age
- `apgar5` = apgar score at 5 minutes

```{r}


lowbwt <- read_xlsx('data/lowbwt.xlsx')
lowbwt


```


Suppose we want to estimate the variance of SBP among low birthweight infants. Here is how we can get both the point estimate and approximate confidence interval.

```{r}

sbp <- lowbwt$sbp

# number of observations
n <- length(sbp)

# sample variance s^2
var_sbp <- var(sbp)

# CI level
alpha <- 0.05

lwr <- (n - 1) * var_sbp / qchisq(1 - alpha/2, df = n - 1)
upr <- (n - 1) * var_sbp / qchisq(alpha/2, df = n - 1)

# R has a package called glue that lets you splice R objects 
# into text using {}. It's very convenient for point and 
# interval summaries

var_sbp <- round(var_sbp, digits = 0)
lwr <- round(lwr, digits = 0)
upr <- round(upr, digits = 0)

glue("the variance (95% CI) of SBP is {var_sbp} ({lwr}, {upr})")

```

Now suppose we want to estimate the proportion of mothers of low birthweight infants who have a toxemia diagnosis during pregnancy. You can get both the point estimate and confidence interval from `binom.test`.

```{r}

# change the 1 in the line below to a 0 if you want
# to compute the proportion without tox
n_tox <- sum(lowbwt$tox == 1)

# note I defined n in the code chunk above.
# changing conf.level to, e.g., 0.90 would give 90% confidence intervals
tox_result <- binom.test(n_tox, n = n, conf.level = 0.95)

# you can see the results here:

tox_result$estimate

tox_result$conf.int

# but whoever reviews your work will appreciate a nicely formatted answer.

est <- round(tox_result$estimate, digits = 2)
lwr <- round(tox_result$conf.int[1], digits = 2)
upr <- round(tox_result$conf.int[2], digits = 2)

glue("The proportion (95% CI) is {est} ({lwr}, {upr})")


```


## Hypothesis Testing

Steps:

1. State the hypotheses (both null and alternative) 
2. Specify the significance level ($\alpha$) 
3. Compute the test statistic 
4. Determine p-value 
5. Compare p-value to the significance level $\alpha$ and decide whether or not to reject the null hypothesis ($H_0$) 
6. State conclusions in terms of subject matter 


### One sample tests

| Test                                      | Null Hypothesis          |
|-------------------------------------------|--------------------------|
| **Mean (μ)**                              |                          |
| &nbsp;&nbsp; One-sample t test (σ not known)           | H₀: μ = μ₀   |
| &nbsp;&nbsp; One-sample z test (σ known)               | H₀: μ = μ₀   |
| **Variance (σ²)**                         |                          |
| &nbsp;&nbsp; One-sample χ² test (only use when data are normal)      | H₀: σ² = σ₀²     |
| **Proportion (p)**                        |                          |
| &nbsp;&nbsp; One-sample normal                         | H₀: p = p₀   |
| &nbsp;&nbsp; One-sample exact                          | H₀: p = p₀   |


Suppose the mean SBP for newborns in the general population is 70 mm Hg. Is the mean SBP for low birthweight infants different than the general population?

```{r}

# One-sample t-test, making use of the sbp vector we made above
t_test_sbp <- t.test(sbp, mu = 70)

# printing the output from the test (all of it)

t_test_sbp

# printing just the results from the test we are interested in

t_test_sbp$estimate

t_test_sbp$conf.int


est <- round(t_test_sbp$estimate, digits = 0)
lwr <- round(t_test_sbp$conf.int[1], digits = 0)
upr <- round(t_test_sbp$conf.int[2], digits = 0)

glue("Mean SBP is {est} (95% CI: {lwr}, {upr})")

# p-value from null hypothesis of mu = 70
t_test_sbp$p.value


```

Suppose that 5% of mothers experience toxemia during pregnancy in the general population. Is the proportion of mothers of low birthweight infants who experience toxemia different than the general population?

```{r}

tox_result <- binom.test(n_tox, n = n, conf.level = 0.95, p = 0.05)

# raw results

tox_result$estimate
tox_result$conf.int
tox_result$p.value


# formatted

est <- round(tox_result$estimate, digits = 2)
lwr <- round(tox_result$conf.int[1], digits = 2)
upr <- round(tox_result$conf.int[2], digits = 2)


glue(
  "The estimated proportion of mothers with toxemia during \\
     pregnancy was {est} (95% CI: {lwr}, {upr}). \\
     Testing whether the estimate is equal to 0.05 yields a p-value \\
     of <0.001"
)


```

### Two sample tests

| Test                                      | Null Hypothesis          |
|-------------------------------------------|--------------------------|
| **Mean (μ) – Paired Samples**            |                          |
| &nbsp;&nbsp; Paired t test               | H₀: Δ = 0                |
| **Mean (μ) – Independent Samples**       |                          |
| &nbsp;&nbsp; Two-sample t test – equal variances | H₀: μ₁ = μ₂       |
| &nbsp;&nbsp; Two-sample t test – unequal variances | H₀: μ₁ = μ₂       |
| **Variance (σ²) – Independent Normal Samples** |                          |
| &nbsp;&nbsp; F test                       | H₀: σ₁² = σ₂²            |


The degree of clinical agreement among physicians on the presence or absence of generalized lymphadenopathy was assessed in 32 randomly selected participants from a prospective study of male sexual contacts of men with acquired immunodeficiency syndrome (AIDS) or an AIDS-related condition (ARC). The total number of palpable lymph nodes was assessed by each of three physicians. Results from two of the three physicians are presented in `lymph_nodes.xlsx`. 

```{r}

lymph_nodes <- read_xlsx("data/lymph_nodes.xlsx")

lymph_nodes

```

The data set has the following variables:

- `ID` = patient ID

- `DocA` = number of palpable lymph nodes found by Doctor A

- `DocB` = number of palpable lymph nodes found by Doctor B

- `Diff` = DocA - DocB

Is there a difference between the assessments of Doctor A vs. Doctor B? Assume the sample size is large enough to rely on the CLT.

```{r}

t_test_diff <- t.test(lymph_nodes$Diff)

# note: this is equivalent to:
t_test_diff <- t.test(x = lymph_nodes$DocA, 
                      y = lymph_nodes$DocB, 
                      paired = TRUE)

t_test_diff

```


Returning back to the `lowbwt` data set, suppose we want to know if SBP differs between males and females.

```{r}

t_test_sbp_by_sex <- with(
  lowbwt,
  t.test(x = sbp[sex == 1],
         y = sbp[sex == 0], 
         paired = FALSE)
)

t_test_sbp_by_sex

```

## Extra practice

Use the “lowbwt.xlsx” data set for these problems.

1. Suppose we are interested in the proportion of low birthweight infants who are female. 


a. Provide a point estimate and 95% confidence interval. 


```{r}

# don't bother with approximations if it's trivial to do the exact test.

binom_test <- binom.test(x = sum(lowbwt$sex==0), n = nrow(lowbwt))

binom_test

```


b. Assume that in the general population, 50% of infants are female. Is the proportion of female infants different for the low birthweight population? Use a hypothesis test to answer this question.

```{r}

# exact test

binom_test <- binom.test(x = sum(lowbwt$sex==0), n = nrow(lowbwt), p = 0.50)

binom_test$p.value

# approximation

p <- mean(lowbwt$sex == 0)
n <- nrow(lowbwt)

mu <- n * p

sigma <- sqrt(n * p * (1 - p))

z <- (mu - 50) / sigma

pval <- 2 * pnorm(z, lower.tail = FALSE)

# Approximation is pretty off here. Another good reason not to do it.
pval

```


2. Does mean SBP differ between infants whose mothers had toxemia during pregnancy and infants whose mothers did not have toxemia? Assume SBP is normally distributed. Use a hypothesis test to answer this question. Also report the 95% confidence interval that corresponds to the test.


```{r}

variances <- lowbwt %>% 
  group_by(tox) %>% 
  summarize(s2 = var(sbp), df = n() - 1)

fstat <- variances$s2[1] / variances$s2[2]

# fstat is >1 so pvalue is 2 * area to the right of fstat

pval_equal_variance <-  2 * pf(q = fstat, 
                               df1 = variances$df[1], 
                               df2 = variances$df[2], 
                               lower.tail = FALSE)

# not enough evidence to conclude variances are not equal
# (note: this is sort of arbitrary - is this sample large enough
#  to detect differences in variance??)

# carry out the t-test with equal variances

t_test_sbp_tox <- with(
  lowbwt,
  t.test(x = sbp[tox == 1],
         y = sbp[tox == 0], 
         paired = FALSE,   # two sample test
         var.equal = TRUE) # equal variances
)

t_test_sbp_tox


```
